Spanner: Google's Globally-Distributed Database

### 2016-01-27 presentation/presentation_spanner.tex
Add visualization of RW transaction

\usetikzlibrary{arrows,positioning,calc}

@@

\institute[PLUS] 
{
Supervisor: Dipl.-Ing. Nikolaus Augsten, PhD \newline \newline \newline
University of Salzburg \\ 
\medskip
Daniel.Kocher@stud.sbg.ac.at
}
\date{\today}

@@

%-------------------------------------------------------------------------------
\section{Motivation}
\begin{frame}
	\frametitle{Motivation}
	\begin{itemize}
    \item{Distributed data at global scale}
    \item{Externally-consistent distributed transactions}
    \item{High availability}
    \item{Two "competitors": \emph{Bigtable} \& \emph{Megastore}}
    \pause
    \item{Why not use Bigtable?}
    \begin{itemize}
      \item{Difficult to use for applications with complex, evolving schemas}
      \item{Only \emph{eventually-consistent} (no strong consistency)}
    \end{itemize}
    \pause
    \item{Why not use Megastore}
    \begin{itemize}
      \item{Poor write throughput}
    \end{itemize}

@@

%-------------------------------------------------------------------------------
\begin{frame}
  \frametitle{Characteristics}
  \begin{center}
    \textbf{Spanner} evolved from a Bigtable-like versioned key-value store to a
    \textbf{temporal multi-version database}
  \end{center}
  \pause
  \begin{itemize}
    \item{Versioned data is stored in schematized semi-relational tables}
    \item{Each version is automatically timestamped with its commit time}
    \item{Garbage Collection \& Read at old timestamps}
    \item{\textbf{General-purpose transactions}}
    \item{SQL-like query language}
  \end{itemize}
\end{frame}

@@

%-------------------------------------------------------------------------------
\subsection{Spanserver Software Stack}
\begin{frame}
  \frametitle{Spanserver Software Stack}
  \only<1>{
    \begin{itemize}
      \item{Responsible for 100 - 1000 \textbf{tablets}}
      \item{A tablet is a bag of key-value mappings \newline
        \texttt{(key:string, timestamp:int64) $\rightarrow$ string}
      }
      \item{Timestamp are assigned to data $\Rightarrow$
        \textbf{multi-version database}
      }
      \item{Tablet states are stored on Colossus (successor of the GFS)}
      \item{Single Paxos state machine on top of each tablet to $\Rightarrow$
        \textbf{Replication}
      }
      \item{Set of replicas is called \textbf{Paxos group}}
    \end{itemize}
  }
  \pause
  \only<2>{
    \begin{itemize}
      \item{Every replica which is a leader implements}
      \begin{itemize}
        \item{a \textbf{lock table} for concurrency control: \texttt{(key range) $\rightarrow$ lock state}}
        \item{a \textbf{transaction manager} (TM) for distributed transactions}
      \end{itemize}
      \item{Transaction manager used to implement \textbf{participant leader}}
      \item{If multiple Paxos groups are involved: Two-Phase Commit}
      \begin{itemize}
        \item{One participant group is chosen as \textbf{coordinator}}
        \item{Participant leader of this group: \textbf{coordinator leader}}
      \end{itemize}
      \item{State of each Transaction manager is stored in Paxos group}
    \end{itemize}
  }
  \pause
  \only<3>{
    \begin{figure}
      \centering
      \includegraphics[width=.7\textwidth]{figures/spanserver-software-stack.png}
      \caption{Spanserver software stack}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Directories \& Placement}
\begin{frame}
  \frametitle{Directories \& Placement}
  \only<1-4>{
    \begin{itemize}
      \item{\textbf{Directories:}}
      \begin{itemize}
        \item{Bucketing abstraction on top of key-value mappings}
        \item{Contain set of contiguous keys sharing common prefix}
        \item{Prefix allows to control locality of data}
        \item{Unit of data placement (same replication configuration)}
        \item{Smallest unit to specify geographic-replication properties (\textbf{placement})}
      \end{itemize}
      \pause
      \item{Data is moved directory-wise between Paxos groups}
      \item{Tablets may have multiple \textbf{partitions} of row space \newline
        $\Rightarrow$ co-locate directories frequently accessed together
      }
      \pause
      \item{Placements can be controlled in two dimensions:}
      \begin{enumerate}
        \item{Number \& types of replicas}
        \item{Geographic placement of replicas}
      \end{enumerate}
      \pause
      \item{\textbf{Movedir:} background task to move directories}
    \end{itemize}
  }
  \only<5>{
    \begin{figure}
      \centering
      \includegraphics[width=.6\textwidth]{figures/directories.png}
      \caption{Directories (the unit of data movement)}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Data Model}
\begin{frame}
  \frametitle{Data Model}
  \only<1>{
    \begin{itemize}
      \item{Features:}
      \begin{itemize}
        \item{Schematized semi-relational tables}
        \item{SQL-like query language}
        \item{General-purpose transactions}
      \end{itemize}
      \item{Layered on top of directory-bucketed key-value mappings}
      \item{Applications create \textbf{databases} in universe}
      \item{Database may contain unlimited number of schematized \textbf{tables}}
      \item{Table contains rows (must have names), columns \& versioned values}
    \end{itemize}
  }
  \only<2>{
    \begin{figure}
      \centering
      \includegraphics[width=.5\textwidth]{figures/interleaving-example.png}
      \caption{Interleaving example}
    \end{figure}
    \begin{itemize}
      \item{Interleaving of tables to form directories is significant}
      \item{Locality relationships between multiple tables $\Rightarrow$ good
        performance
      }
    \end{itemize}
  }
\end{frame}

%-------------------------------------------------------------------------------
\section{TrueTime}
\begin{frame}
  \frametitle{TrueTime}
  \only<1>{
    \begin{table}
      \centering
      \begin{tabular}{|l||l|}
        \hline
        Method & Returns \tabularnewline
        \hline\hline
        $TT.now()$ & $TTinterval$: [$earliest$, $latest$] \tabularnewline
        \hline
        $TT.after(t)$ & true if $t$ has definitely passed \tabularnewline
        \hline
        $TT.before(t)$ & true if $t$ has definitely not arrived \tabularnewline
        \hline
      \end{tabular}
      \caption{TrueTime API ($t$ is of type $TTstamp$)}
    \end{table}
    Time is represented as a $TTinterval$ (interval with bounded uncertainty)
    \newline\newline
    Let $e$ be an event and $t_{abs}(e)$ denote the absolute time of $e$.
    \newline
    For $tt = TT.now()$, the following property holds:
    \begin{center}
      $tt.earliest \leq t_{abs}(e) \leq tt.latest$
    \end{center}
  }
  \only<2>{
    \begin{figure}
      \centering
      \begin{tikzpicture}[thick]
        \node (left-end) at (0, 0) {};
        \node[right = 6 of left-end] (right-end) {};

@@

        \draw[<->] (left-bound-e.center) -- (right-bound-e.center);
        %\node[below right= 1 and 2.5 of left] {$2\epsilon$};
      \end{tikzpicture}
      \caption{Visualization of $TT.now()$ and $\epsilon$}
    \end{figure}
  }
  \only<3>{
    \begin{itemize}
      \item{Time references: GPS and atomic clocks (different failure modes)}
      \item{Set of \textbf{time master} machines per datacenter}
      \item{\textbf{Armageddon masters}: masters using atomic clocks}
      \item{\textbf{timeslave daemon} per machine}
      \pause
      \item{Masters compare their time references regularly}
      \item{Daemons poll a variety of masters (liar detection)}
      \pause
      \item{$\epsilon$ is the instantaneous error bound (typically 1 - 7ms)}
      \item{$\overline{\epsilon}$ is the average error bound (typically 4ms)}
      \item{$\epsilon$ is derived from applied worst-case local clock drift ($200\mu s/sec$)}
    \end{itemize}
  }
\end{frame}

@@

%-------------------------------------------------------------------------------
\subsection{Timestamp Management}
\begin{frame}
  \frametitle{Timestamp Management}
  \only<1>{
    \footnotesize
    \begin{table}
      \begin{tabular}{|l||l|l|}
        \hline
        Operation & CC & Replica required\tabularnewline
        \hline\hline
        Read-Write transaction & Pessimistic & leader\tabularnewline
        \hline
        Read-Only transaction & Lock-free & leader; any
        \tabularnewline
        \hline
        Snapshot Read, client-provided timestamp & Lock-free & any\tabularnewline
        \hline
        Snapshot Read, client-provided bound & Lock-free & any \tabularnewline
        \hline
      \end{tabular}
      \caption{Supported types of reads \& writes}
    \end{table}
  }
  \begin{itemize}
    \item{Standalone write $\Rightarrow$ RW transaction}
    \item{Non-snapshot standalone read $\Rightarrow$ RO transaction}
    \item{RO transactions}
    \begin{itemize}
      \item{must predeclare to not include writes}
      \item{can proceed on any sufficiently up-to-date-replica (also snapshot reads)}
    \end{itemize}
  \end{itemize}
\end{frame}

@@

%-------------------------------------------------------------------------------
\subsubsection{Timestamps in RW Transactions}
\begin{frame}
  \frametitle{Timestamps in RW Transactions}
  \begin{itemize}
    \item{Timestamp assignment only in between the two phases (2PL)}
    \item{\textbf{Monotonicity variant:} \newline
      Within each Paxos group, Spanner assigns timestamps to Paxos writes in
      monotonically increasing order, even across leaders
    }
    \item{Enforced across leaders by disjointness invariant: \newline
      Leader must only assign timestamps within the interval of leader lease
    }
    \item{Timestamp $s$ assigned $\Rightarrow$ $s_{max} = s$}
    \item{\textbf{External consistency invariant}: \newline
      \begin{center}
        $t_{abs}\left(e^{commit}_1\right) < t_{abs}\left(e^{start}_2\right)$
        $\Rightarrow s_1 < s_2$
      \end{center}
    }
  \end{itemize}
\end{frame}

@@

\subsubsection{Read-Write Transactions}
\begin{frame}
  \frametitle{Read-Write Transactions}
  \begin{itemize}
    \item{RW transactions buffered at client until Commit}
    \item{Reads do not see effects of transaction's writes}
    \item{Client has completed all reads \& buffered all writes $\Rightarrow$ 2PC}
    \pause
    \item{Non-coordinator-participant leader}
    \begin{enumerate}
      %\item{Acquires write locks}
      \item{Chooses prepare timestamp according to monotonicity}
      \item{Logs prepare record through Paxos}
      \item{Notifies coordinator of its prepare timestamp}
    \end{enumerate}
    \pause
    \item{Coordinator leader}
    \begin{enumerate}
      %\item{Aquires write locks}
      \item{Chooses timestamp for whole transaction}
      \item{Commit timestamp $s$ must be}
      \begin{itemize}
        \item{$\geq$ all prepare timestamps}
        \item{$> TT.now().latest$ at the time the Commits were received}
        \item{$>$ any previously assigned timestamp}
      \end{itemize}
      \item{Logs Commit record through Paxos}
    \end{enumerate}
  \end{itemize}
\end{frame}

>>>>

\usetikzlibrary{arrows,positioning,calc,shapes}

@@

\institute[PLUS] 
{
Summarizing Talk \newline
Efficent Algorithms Seminar \newline
University of Salzburg
}
\date{\today}

@@

%-------------------------------------------------------------------------------
\section{Motivation}
\begin{frame}
	\frametitle{Motivation}
	\begin{itemize}
    \item{Distributed data at global scale}
    \item{Externally-consistent distributed transactions}
    \item{High availability}
    \item{Two "competitors": \emph{Bigtable} \& \emph{Megastore}}
    \pause
    \item{Why not use Bigtable?}
    \begin{itemize}
      \item{Difficult to use for applications with complex, evolving schemas}
      \item{Only \textbf{eventually-consistent} (no strong consistency)}
    \end{itemize}
    \pause
    \item{Why not use Megastore}
    \begin{itemize}
      \item{Poor write throughput}
    \end{itemize}

@@

%-------------------------------------------------------------------------------
\begin{frame}
  \frametitle{Characteristics}
  \begin{center}
    \textbf{Spanner} evolved from a Bigtable-like versioned key-value store to a
    \textbf{temporal multi-version database}
  \end{center}
  \pause
  \begin{itemize}
    \item{Versioned data is stored in schematized semi-relational tables}
    \item{Each version is automatically timestamped with its commit time}
    \item{Garbage collection \& reads at old timestamps}
    \item{\textbf{General-purpose transactions}}
    \item{SQL-like query language}
  \end{itemize}
\end{frame}

@@

%-------------------------------------------------------------------------------
\subsection{Spanserver Software Stack}
\begin{frame}
  \frametitle{Spanserver Software Stack}
  \only<1>{
    \begin{itemize}
      \item{Responsible for 100 - 1000 \textbf{tablets}}
      \item{A tablet is a bag of key-value mappings \newline
        \texttt{(key:string, timestamp:int64) $\rightarrow$ string}
      }
      \item{Timestamps are assigned to data $\Rightarrow$
        \textbf{multi-version database}
      }
      \item{Tablet states are stored on Colossus}
      \item{Single Paxos state machine on top of each tablet $\Rightarrow$
        \textbf{Replication}
      }
      \item{Set of replicas is called \textbf{Paxos group}}
    \end{itemize}
  }
  \pause
  \only<2>{
    \begin{itemize}
      \item{Every replica which is a leader implements}
      \begin{itemize}
        \item{a \textbf{lock table} for concurrency control: \texttt{(key range) $\rightarrow$ lock state}}
        \item{a \textbf{transaction manager} (TM) for distributed transactions}
      \end{itemize}
      \item{Transaction manager used to implement \textbf{participant leader}}
      \item{If multiple Paxos groups are involved: Two-Phase Commit}
      \begin{itemize}
        \item{One participant group is chosen as \textbf{coordinator}}
        \item{Participant leader of this group: \textbf{coordinator leader}}
      \end{itemize}
      \item{State of each transaction manager is stored in Paxos group}
    \end{itemize}
  }
  \pause
  \only<3>{
    \begin{figure}
      \centering
      \includegraphics[width=.7\textwidth]{figures/spanserver-software-stack.png}
      \caption{Spanserver software stack}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Directories \& Placement}
\begin{frame}
  \frametitle{Directories \& Placement}
  \only<1-3>{
    \begin{itemize}
      \item{\textbf{Directories:}}
      \begin{itemize}
        \item{Bucketing abstraction on top of key-value mappings}
        \item{Contain set of contiguous keys sharing common prefix}
        \item{Control locality of data}
        \item{Unit of data placement (same replication configuration)}
        \item{Smallest unit to specify geographic-replication properties (\textbf{placement})}
      \end{itemize}
      \pause
      \item{Data is moved directory-wise between Paxos groups}
      \item{Tablets may have multiple \textbf{partitions} of row space \newline
        $\Rightarrow$ co-locate directories frequently accessed together
      }
      \pause
      \item{Placements can be controlled in two dimensions:}
      \begin{enumerate}
        \item{Number \& types of replicas}
        \item{Geographic placement of replicas}
      \end{enumerate}
      \item{\textbf{Movedir:} background task to move directories}
    \end{itemize}
  }
  \only<4>{
    \begin{figure}
      \centering
      \includegraphics[width=.6\textwidth]{figures/directories.png}
      \caption{Directories (the unit of data movement)}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Data Model}
\begin{frame}
  \frametitle{Data Model}
  \only<1>{
    \begin{itemize}
      \item{Features:}
      \begin{itemize}
        \item{Schematized semi-relational tables}
        \item{SQL-like query language}
        \item{General-purpose transactions}
      \end{itemize}
      \item{Layered on top of directory-bucketed key-value mappings}
      \item{Applications create \textbf{databases} in universe}
      \item{Database may contain unlimited number of schematized \textbf{tables}}
      \item{Table contains rows (must have names), columns \& versioned values}
    \end{itemize}
  }
  \only<2>{
    \begin{itemize}
      \item{Databases must be partitioned in one or more hierarchies}
      \item{\textbf{Directory table}: table at the top}
      \item{Directory contains ordered rows starting at a directory table key}
      \item{Interleaving tables to form directories is significant}
      \item{Locality relationships between multiple tables $\Rightarrow$ good
        performance
      }
    \end{itemize}

    \begin{figure}
      \centering
      \includegraphics[width=.5\textwidth]{figures/interleaving-example.png}
      \caption{Interleaving example}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\section{TrueTime}
\begin{frame}
  \frametitle{TrueTime}
  \only<1>{
    \begin{table}
      \centering
      \begin{tabular}{|l||l|}
        \hline
        Method & Returns \tabularnewline
        \hline\hline
        $TT.now()$ & $TTinterval$: [$earliest$, $latest$] \tabularnewline
        \hline
        $TT.after(t)$ & true if $t$ has definitely passed \tabularnewline
        \hline
        $TT.before(t)$ & true if $t$ has definitely not arrived \tabularnewline
        \hline
      \end{tabular}
      \caption{TrueTime API ($t$ is of type $TTstamp$)}
    \end{table}
    Time is represented as a $TTinterval$ (interval with bounded uncertainty)
    \newline\newline
    Let $e$ be an event and $t_{abs}(e)$ denote the absolute time of $e$.
    \newline
    For $tt = TT.now()$, the following property holds:
    \begin{center}
      $tt.earliest \leq t_{abs}(e) \leq tt.latest$
    \end{center}
  }
  \only<2>{
    \begin{itemize}
      \item{Time references: GPS and atomic clocks}
      \item{Set of \textbf{time master} machines per datacenter}
      \item{\textbf{Armageddon masters}: masters using atomic clocks}
      \item{\textbf{Timeslave daemon} on each machine}
      \pause
      \item{Masters compare their time references regularly}
      \item{Daemons poll a variety of masters (liar detection)}
      \pause
      \item{$\epsilon$ is the instantaneous error bound (typically 1 - 7ms)}
      \item{$\overline{\epsilon}$ is the average error bound (typically 4ms)}
      \item{$\epsilon$ is derived from applied worst-case local clock drift ($200\mu s/sec$)}
    \end{itemize}
  }
  \only<3>{
    \begin{figure}
      \centering
      \begin{tikzpicture}[thick]
        \node (left-end) at (0, 0) {};
        \node[right = 6 of left-end] (right-end) {};

@@

        \draw[<->] (left-bound-e.center) -- (right-bound-e.center);
      \end{tikzpicture}
      \caption{Visualization of $TT.now()$ and $\epsilon$}
    \end{figure}
  }
\end{frame}

@@

%-------------------------------------------------------------------------------
\subsection{Timestamp Management}
\begin{frame}
  \frametitle{Timestamp Management}
  \only<1>{
    \footnotesize
    \begin{table}
      \begin{tabular}{|l||l|l|}
        \hline
        Operation & CC & Replica required\tabularnewline
        \hline\hline
        Read-Write transaction & Pessimistic & leader\tabularnewline
        \hline
        Read-Only transaction & Lock-free & leader; any
        \tabularnewline
        \hline
        Snapshot Read, client-provided timestamp & Lock-free & any\tabularnewline
        \hline
        Snapshot Read, client-provided bound & Lock-free & any \tabularnewline
        \hline
      \end{tabular}
      \caption{Supported types of reads \& writes}
    \end{table}
  }
  \begin{itemize}
    \item{Standalone write $\Rightarrow$ RW transaction}
    \item{Non-snapshot standalone read $\Rightarrow$ RO transaction}
    \item{RO transactions}
    \begin{itemize}
      \item{must predeclare to not include writes}
      \item{can proceed on any sufficiently up-to-date replica (also snapshot
        reads)
      }
    \end{itemize}
  \end{itemize}
\end{frame}

@@

%-------------------------------------------------------------------------------
\subsubsection{Timestamps in RW Transactions}
\begin{frame}
  \frametitle{Timestamps in RW Transactions}
  \only<1>{
    \begin{itemize}
      \item{Timestamp assignment only in between the two phases (2PL)}
      \item{\textbf{Monotonicity variant:} \newline
        Within each Paxos group, Spanner assigns timestamps to Paxos writes in
        monotonically increasing order, even across leaders
      }
      \item{Enforced across leaders by disjointness invariant: \newline
        Leader must only assign timestamps within the interval of leader lease
      }
      \item{Timestamp $s$ assigned $\Rightarrow$ $s_{max} = s$}
      \item{\textbf{External consistency invariant}: \newline
        \begin{center}
          $t_{abs}\left(e^{commit}_1\right) < t_{abs}\left(e^{start}_2\right)$
          $\Rightarrow s_1 < s_2$
        \end{center}
      }
    \end{itemize}
  }
  \only<2>{
    \begin{itemize}
      \item{Let $e^{server}_i$ denote the arrival event of commit request for
        $T_i$
      }
      \item{Two rules guarantee the external consistency invariant:}
      \begin{enumerate}
        \item{\textbf{Start} \newline
          Coordinator leader assigns commit timestamp \newline
          $s_i \geq TT.now().latest$ after $e^{server}_i$
        }
        \item{\textbf{Commit Wait} \newline
          Coordinator leader ensures clients cannot see \newline
          data commited by $T_i$ until $TT.after(s_i)$ is true
        }
      \end{enumerate}
    \end{itemize}
  }
\end{frame}

@@

\subsubsection{Read-Write Transactions}
\begin{frame}
  \frametitle{Read-Write Transactions}
  \only<1-3>{
    \begin{itemize}
      \item{RW transactions buffered at client until Commit}
      \item{Reads do not see effects of transaction's writes}
      \item{Client has completed all reads \& buffered all writes $\Rightarrow$ 2PC}
      \pause
      \item{Non-coordinator-participant leader}
      \begin{enumerate}
        %\item{Acquires write locks}
        \item{Chooses prepare timestamp according to monotonicity}
        \item{Logs prepare record through Paxos}
        \item{Notifies coordinator of its prepare timestamp}
      \end{enumerate}
      \pause
      \item{Coordinator leader}
      \begin{enumerate}
        %\item{Aquires write locks}
        \item{Chooses timestamp for whole transaction}
        \item{Commit timestamp $s$ must be}
        \begin{itemize}
          \item{$\geq$ all prepare timestamps}
          \item{$> TT.now().latest$ at the time the Commits were received}
          \item{$>$ any previously assigned timestamp}
        \end{itemize}
        \item{Logs Commit record through Paxos}
      \end{enumerate}
    \end{itemize}
  }
  \only<4-9>{
    \begin{figure}
      \centering
      \begin{tikzpicture}[thick]
        \node (left-end-t1) at (0,2) {};
        \node[right = 9 of left-end-t1] (right-end-t1) {};
        \draw[ultra thick, ->, >=stealth] (left-end-t1.center) -- (right-end-t1.center)
          node[right] (right-end-t1-t) {$t$};
          \node[left = 0 of left-end-t1.west] (left-end-t1-t) {$\mathbf{T_C}$};

        \node[below = 1 of left-end-t1] (left-end-t2) {};
        \node[right = 9 of left-end-t2] (right-end-t2) {};
        \draw[->, >=stealth] (left-end-t2.center) -- (right-end-t2.center)
          node[right] (right-end-t2-t) {$t$};
        \node[left = 0 of left-end-t2.west] (left-end-t2-t) {$T_2$};

        \node[below = 1 of left-end-t2] (left-end-t3) {};
        \node[right = 9 of left-end-t3] (right-end-t3) {};
        \draw[->, >=stealth] (left-end-t3.center) -- (right-end-t3.center)
          node[right] (right-end-t3-t) {$t$};
        \node[left = 0 of left-end-t3.west] (left-end-t3-t) {$T_3$};

        \node[right = 0.25 of left-end-t1] (acquire-locks-t1) {};
        \node[right = 1.05 of left-end-t2] (acquire-locks-t2) {};
        \node[right = 0.65 of left-end-t3] (acquire-locks-t3) {};
        \foreach \x in {1, 2, 3}{
          \draw[blue] (acquire-locks-t\x.center) -- ++(0, -0.25);
          \draw[blue] (acquire-locks-t\x.center) -- ++(0, 0.25);
        }

        \node[above = 0.75 of acquire-locks-t1]
          (acquire-locks-t) {\scriptsize\color{blue}Acquire all locks};
        \

        \only<5-9>{
          \foreach \x in {1, 2, 3}{
            \node[right = -0.1 of acquire-locks-t\x] (compute-s-t\x) {};
            %\draw (compute-s-t\x.center) -- ++(0, -0.15);
            %\draw (compute-s-t\x.center) -- ++(0, 0.15);
          }
          \node[below = 1 of acquire-locks-t3.center]
            (compute-s-t) {\scriptsize $s=TT.now()$};
          \draw[<-, >=stealth, dotted] (compute-s-t1.center) -- ++(0, -3.5);
          \draw[<-, >=stealth, dotted] (compute-s-t2.center) -- ++(0, -2.25);
          \draw[<-, >=stealth, dotted] (compute-s-t3.center) -- ++(0, -0.9);
        }
        \only<6-9>{
          \node[right = 1 of compute-s-t2] (start-log-t2) {};
          \node[right = 1.75 of compute-s-t3] (start-log-t3) {};
          \node[above = 2 of start-log-t2] (start-log-t)
            {\scriptsize Start Logging};
          \draw[<-, >=stealth, dotted] (start-log-t2.center) -- ++(0, 2.25);
          \draw[<-, >=stealth, dotted] (start-log-t3.center) -- ++(0, 3.5);
        }
        \only<7-9>{
          \node[right = 1 of start-log-t3] (end-log-t3) {};
          \node[right = 1.75 of start-log-t2] (end-log-t2) {};
          \node[above = 2 of end-log-t2] (end-log-t)
            {\scriptsize Logging done};
          \draw[<-, >=stealth, dotted] (end-log-t2.center) -- ++(0, 2.25);
          \draw[<-, >=stealth, dotted] (end-log-t3.center) -- ++(0, 3.5);
          \node[right = 5.5 of compute-s-t1] (compute-s-global) {};
          \draw[<-, >=stealth, black!50!green] (compute-s-global) ++(-0.1, 0) --
            (end-log-t3.north east);
          \draw[<-, >=stealth, black!50!green] (compute-s-global) ++(-0.65, 0) --
            (end-log-t2.north east);
          \draw[black!50!green, ultra thick] (compute-s-global.center) -- ++(0, -0.25);
          \draw[black!50!green, ultra thick] (compute-s-global.center) -- ++(0, 0.25);

          \node[above = 0.25 of compute-s-global] (compute-s-global-t)
            {\scriptsize\color{black!50!green} Overall $s$};
          \node[below = 0 of end-log-t3] (prepared-t)
            {\scriptsize\color{black!50!green}Prepared; Send $s$};
        }
        \only<8-9>{
          \node[right = 0.5 of compute-s-global] (commit-wait-done) {};
          \node[below = 3.45 of commit-wait-done] (commit-wait-done-t)
            {\scriptsize Commit Wait done};
          \draw[<-, >=stealth, dotted] (commit-wait-done.center) -- ++(0, -3.5);
          \node[right = 0 of commit-wait-done] (release-locks-t1) {};
          \draw[red] (release-locks-t1.center) -- ++(0, -0.25);
          \draw[red] (release-locks-t1.center) -- ++(0, 0.25);
          \node[above right = 0.75 and -0.75 of release-locks-t1]
            (release-locks-t1-t) {\scriptsize\color{red}Release all locks};
        }
        \only<9>{
          \node[right = -0.15 of release-locks-t1] (notify-t1) {};
          \node[right = 3.1 of end-log-t2] (notify-t2) {};
          \node[right = 3.35 of end-log-t3] (notify-t3) {};
          \foreach \x in {2, 3}{
            \draw[->, >=stealth, black!50!green] (notify-t1.center) --
              (notify-t\x.center); 
          }
          %\node[below = 0 of notify-t3] (notify-t)
          %  {\scriptsize\color{black!50!green}Notify participants};
          \node[right = -0.15 of notify-t2] (release-locks-t2) {};
          \node[right = -0.15 of notify-t3] (release-locks-t3) {};
          \draw[red] (release-locks-t2.center) -- ++(0, -0.25);
          \draw[red] (release-locks-t2.center) -- ++(0, 0.25);
          \draw[red] (release-locks-t3.center) -- ++(0, -0.25);
          \draw[red] (release-locks-t3.center) -- ++(0, 0.25);
        }
     \end{tikzpicture}
      \caption{RW transaction illustration}
    \end{figure}
  }
\end{frame}
